---
title: "Evaluations of positive and negative stimuli using the Affective Misattribution Procedure (AMP) and self-reports"
subtitle: "Data processing"
<<<<<<< HEAD
author: "Template: Ian Hussey; content: [Student name]"
=======
author: "Template: Ian Hussey; content: [Lukas Müller]"
>>>>>>> 992b80d467dcb21559c009b9127196cba2aaf2f1
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

```

# Dependencies

```{r}

<<<<<<< HEAD
# install.packages("dplyr")

=======
>>>>>>> 992b80d467dcb21559c009b9127196cba2aaf2f1
library(dplyr)
library(tidyr)
library(readr)
library(janitor) # for clean_names()
library(stringr)

```

# Get data

```{r}

# demographics
data_demographics_raw_messy <- read_csv("../data/raw/data_demographics_raw_messy.csv", skip = 2) |>
  janitor::clean_names()

data_demographics_raw <- read_csv("../data/raw/data_demographics_raw.csv") |>
  janitor::clean_names()

# self report measure
data_selfreport_raw <- read_csv("../data/raw/data_selfreport_raw.csv") |>
  janitor::clean_names()

# affect attribution procedure
data_amp_raw <- read_csv("../data/raw/data_amp_raw.csv") |>
  janitor::clean_names()

```

# Demographics

Clean the demographics data so that summaries of age and gender can be made in the analysis.

```{r}

<<<<<<< HEAD
# assignment
# <- 
# x <- y
# = 
# x = y

# explanation of the pipe(s)

# %>% # tidyverse
  
# |> # base R version
  
# dat_age_gender <- data_demographics_raw %>%
#   rename(date_time = date) %>%
#   mutate(group = group + 1)  

# dat_age_gender <- data_demographics_raw |>
#   rename(date_time = date) |>
#   mutate(group = group + 1) 

# colnames(dat_age_gender)

# dat_age_gender <- 
#   select(data_demographics_raw, subject, trialcode, response)
# 
# dat_age_gender <- 
#   select(.data = data_demographics_raw, subject, trialcode, response)

# dat_age_gender <- data_demographics_raw |>
#   rename(boobleborb = subject, item = trialcode) |>
#   select(boobleborb, item, response)

# dat_age_gender <- data_demographics_raw |>
#   select(subject, trialcode, response) |>
#   rename(unique_id = subject, item = trialcode)
# 
# temp <- dat_age_gender |>
#   filter(item == "age")

# # wont work, as age isn't a defined object
# temp <- dat_age_gender |>
#   filter(item == age)

# # there are much more flexible ways of doing 'string matching'
# temp <- dat_age_gender |>
#   filter(item %in% c("age", "Age", "AGE")) # any item of this vector
# 
# temp <- dat_age_gender |>
#   filter(item == "age" | 
#            item == "Age") # any of these
# 
# # | # or operator
# # & # and operator
# 
# library(stringr)
# temp <- dat_age_gender |>
#   filter(str_detect(item, "age"))

# # wrangle the demographics data out of the raw
# dat_age_gender <- data_demographics_raw |>
#   # rename for clarity
#   rename(unique_id = subject, item = trialcode) |>
#   # select useful columns
#   select(unique_id, item, response) |>
#   # retain rows that we need
#   filter(item == "age")

# temp <- dat_age_gender |>
#   distinct(response)
# 
# 
# # check data type for each column
# sapply(dat_age_gender, class)


# temp2 <- dat_age_gender |>
#   # mutate allows you to create or change variables
#   mutate(unique_id = unique_id * 2)


library(stringr)

# # more generic, but puts more trust in your understanding chatGPT
# temp2 <- dat_age_gender |>
#   mutate(response = str_remove_all(response, "\\D"),
#          response = na_if(response, ""))
# 
# 
# str_remove_all("twenty", "\\D")
# str_remove_all("22.5", "\\D")


# # more manual
# temp2 <- dat_age_gender |>
#   mutate(response = ifelse(response %in% c("old", "lol", ""), NA, response))


# wrangle the demographics data out of the raw
dat_age_gender <- data_demographics_raw |>
  # rename for clarity
  rename(unique_id = subject, item = trialcode) |>
  # select useful columns
  select(unique_id, item, response) |>
  # retain rows that we need
  filter(item == "age") |>
  # remove non-numbers or decimals from age
  mutate(response = str_remove_all(response, "[^\\d.]"),  # retains only digits and decimal points
         response = na_if(response, ""),
         response = as.numeric(response))

```

=======
dat_age_gender <- data_demographics_raw %>% 
  select(trialcode, response)

```

```{r}
gender_data_raw <- dat_age_gender %>%
  filter(trialcode == "gender") %>%
  select(response) %>%
  rename(gender = response)

head(gender_data)
```

```{r}
gender_data_clean <- gender_data_raw %>% 
  mutate(gender = ifelse(is.na(as.numeric(gender)), gender, NA))

head(gender_data_clean)
```

```{r}

gender_data_cleaner <- gender_data_clean %>%
  mutate(gender = case_when(
    gender %in% c("Male", "male", "MALE") ~ 1,
    gender %in% c("Female", "female", "FEMALE") ~ 2,
    TRUE ~ as.numeric(gender)))

head(gender_data_cleaner)

```

```{r}

gender_summary <- gender_data_cleaner %>%
  summarise(
    count_1 = sum(gender == 1, na.rm = TRUE),
    count_2 = sum(gender == 2, na.rm = TRUE),
    count_NA = sum(is.na(gender)),
    count_total = sum(is.na(gender) + gender == 1 | 2))


gender_summary

```

```{r}

gender_summary <- data_demographics_raw %>% 
  select(trialcode, response) %>%
  filter(trialcode == "gender") %>%
  select(response) %>%
  rename(gender = response) %>%
  mutate(gender = ifelse(is.na(as.numeric(gender)), gender, NA)) %>%
  mutate(gender = ifelse(gender %in% c("Male", "male", "MALE"), 1,
                         ifelse(gender %in% c("Female", "female", "FEMALE"), 2, NA))) %>%
  summarise(
    count_1 = sum(gender == 1, na.rm = TRUE),
    count_2 = sum(gender == 2, na.rm = TRUE),
    count_NA = sum(is.na(gender)),
    count_total = sum(gender %in% c(1, 2), na.rm = TRUE)
  )


head(gender_summary)
                
                         
```


>>>>>>> 992b80d467dcb21559c009b9127196cba2aaf2f1
# Exclusions / data quality

## AMP

Create an exclusion variable `exclude_amp_performance` based on AMP performance: "exclude" if more than 10% of trials are < 100ms, or "include" if not.

Create an exclusion variable `exclude_amp_completeness` based on the AMP having the right number of trials.

```{r}

#data_amp_performance_criteria <- data_amp_raw 

```

# Self-reports

Create a mean score for each participant of the three self-report items.

```{r}

#data_selfreport_trial_level <- data_selfreport_raw 

```

# Affect Misattribution Procedure

Create an overall AMP bias score. Score each trial as having been evalauted as congruent with the prime or not. eg When trialcode == "prime_positive" & evaluative_response == 1, then record as 1, etc. AMP score is the proportion of congruent responses.

```{r}

#data_amp_score_congruence <- data_amp_raw 

```

# Combine 

Combine data frames together to create one wide-format data frame where each row represents a participant.

Flag all duplicates for exclusion.

```{r}

# data_processed_before_exclusions 

```

# Define master exclusion variable

```{r}

#data_processed <- data_processed_before_exclusions

```

# Write to disk

```{r}

# # in case this dir doesn't exist, create it
# dir.create("../data/processed/")
# 
# # save data to disk in that dir
# write_csv(data_processed, "../data/processed/data_processed.csv")

```

# Session info

```{r}

sessionInfo()

```


