---
title: "Demonstration of collider bias as a form of model misspecification"
author: "Ian Hussey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}

# set default chunk options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

# dependencies
library(tidyr)
library(dplyr)
library(purrr) 
library(lavaan)
library(semPlot)
library(knitr)
library(kableExtra)
library(parameters)

```

# Collider population analyzed with independent effects (multiple regression) analysis

## Plots

Data generating process vs. model used in analysis

```{r fig.height=6, fig.width=7}

# collider
mod_collider <- "M ~ X + Y; Y ~ X"

layout_matrix <- matrix(c( 0,  1,
                           1,  0,
                          -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_collider, sample.nobs = 100) %>%
  sem(model = mod_collider, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Data generating process in the population: Collider", adj = 0)



# independent effects (multiple regression)
mod_independenteffects    <- "Y ~ X + M" 

layout_matrix <- matrix(c( 1,  0,
                          -1,  0,
                           0,  1), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_independenteffects, sample.nobs = 100) %>%
  sem(model = mod_independenteffects, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Model used in analysis: independent effects (multiple regression)", adj = 0)

```

## Simulate

```{r}

# define data generating process ----
generate_data <- function(population_model, n) {
  
  data <- lavaan::simulateData(model = population_model, sample.nobs = n) |>
    mutate(id = row_number())
    
  return(data)
}

# define analysis ----
# extract the regression estimate for y ~ x
# nb requires that analysis_model includes a y ~ x effect
analyse <- function(data, analysis_model){

  # fit model
  fit <- sem(model = analysis_model, data = data)
  
  # extract results
  results <- fit |>
    parameters() |>
    as_tibble() |>
    filter(To == "Y" & From == "X") |>
    dplyr::select(estimate_yx = Coefficient, p_yx = p)
  
  return(results)
}

# run simulation ----
simulation <- 
  # define simulation conditions
  tibble(n = 200,
         population_model = "M ~ 0.5*X + 0.5*Y; Y ~~ 0.0*X",
         population_label = "collider",
         population_yx = 0,
         analysis_model = "Y ~ X + M", 
         analysis_label = "independent effects (multiple regression)") |>
  expand_grid(iteration = 1:1000) |>
  # for each one, generate data
  mutate(data = pmap(list(population_model, n), generate_data)) |>
  # for each dataset, analyze it
  mutate(results = pmap(list(data, analysis_model), analyse)) 

# summarize across iterations ----
simulation_summary <- simulation |>
  unnest(results) |>
  group_by(population_label, analysis_label, population_yx) |>
  summarize(mean_estimate_yx = mean(estimate_yx), 
            proportion_significant = mean(p_yx < .05),
            .groups = "drop")

# print table of results
simulation_summary |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Collider population analyzed with mediation analysis

## Plots

Data generating process vs. model used in analysis

```{r fig.height=6, fig.width=7}

# collider
mod_collider <- "M ~ X + Y; Y ~ X"

layout_matrix <- matrix(c( 0,  1,
                           1,  0,
                          -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_collider, sample.nobs = 100) %>%
  sem(model = mod_collider, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Data generating process in the population: Collider", adj = 0)



# mediation
mod_mediation <- "Y ~ M + X; M ~ X"

layout_matrix <- matrix(c( 1,  0,
                           0,  1,
                          -1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_mediation, sample.nobs = 100) %>%
  sem(model = mod_mediation, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Model used in analysis: mediation", adj = 0)



# mediation reversed
mod_mediation <- "X ~ M + Y; M ~ Y"

layout_matrix <- matrix(c(-1,  0,
                           0,  1,
                           1,  0), 
                        ncol = 2, 
                        byrow = TRUE)

lavaan::simulateData(model = mod_mediation, sample.nobs = 100) %>%
  sem(model = mod_mediation, data = .) |>
  semPaths(whatLabels = "diagram", 
           layout = layout_matrix, 
           residuals = FALSE,
           edge.label.cex = 1.2, 
           sizeMan = 10)
title(main = "Other possible models: reverse mediation", adj = 0)

```

## Simulate

```{r}

# define data generating process ----
generate_data <- function(population_model, n) {
  
  data <- lavaan::simulateData(model = population_model, sample.nobs = n) |>
    mutate(id = row_number())
    
  return(data)
}

# define analysis ----
# extract the regression estimate for y ~ x
# nb requires that analysis_model includes a y ~ x effect
analyse <- function(data, analysis_model){

  # fit model
  fit <- sem(model = analysis_model, data = data)
  
  # extract results
  results <- fit |>
    parameters() |>
    as_tibble() |>
    filter(To == "Y" & From == "X") |>
    dplyr::select(estimate_yx = Coefficient, p_yx = p)
  
  return(results)
}

# run simulation ----
simulation <- 
  # define simulation conditions
  tibble(n = 200,
         population_model = "M ~ 0.5*X + 0.5*Y; Y ~~ 0.0*X",
         population_label = "collider",
         population_yx = 0,
         analysis_model = "Y ~ M + X; M ~ X", 
         analysis_label = "mediation") |>
  expand_grid(iteration = 1:1000) |>
  # for each one, generate data
  mutate(data = pmap(list(population_model, n), generate_data)) |>
  # for each dataset, analyze it
  mutate(results = pmap(list(data, analysis_model), analyse)) 

# summarize across iterations ----
simulation_summary <- simulation |>
  unnest(results) |>
  group_by(population_label, analysis_label, population_yx) |>
  summarize(mean_estimate_yx = mean(estimate_yx), 
            proportion_significant = mean(p_yx < .05),
            .groups = "drop")

# print table of results
simulation_summary |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 2) |>
  kable() |>
  kable_classic(full_width = FALSE)

```

# Extensions

- What more complex multivariable regression models do you use in your research or do you observe in the literature? 
- What is the causal justification for their structure? [hard question]
- Try to implement such a model in lavaan syntax to generate data for it. For example, add other continuous covariates such as age or SES or other psycholgical variables. 
- What impact does changing the causal structure among them have on the results you observe? [hard question]

# Further reading

Craig Seawall has an excellent [twitter thread](https://x.com/cjsewall9/status/1782398602071114047) and [article](https://doi.org/10.1037/abn0000950) on how an over simplified and poorly specified theory can generate spurious and artifact results. For example, a poor understanding of the broader causal structure can make it appear as if social media and gender interact to cause poor mental health, when this is not the case.

# Session info

```{r}

sessionInfo()

```


