---
title: "Reshaping data with {tidyr}' pivot_wider() and pivot_longer()"
author: "Template: Ian Hussey; Content: [student name]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

# Resources

See code and gifs [here](https://github.com/gadenbuie/tidyexplain?tab=readme-ov-file#pivot-wider-and-longer) which illustrate pivots (and indeed other tidyverse verbs).

```{r, include=FALSE}

# set knit options
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

dir.create("plots")

```

pivots helfen uns das format der daten zu ändern

# Dependencies

```{r}

library(dplyr)
library(tidyr)
library(tibble)
install.packages("devtools")
devtools::install_github("debruine/faux")
library(faux)
library(janitor)
library(ggplot2)
library(scales)
library(psych)
library(readr)
library(knitr)
library(kableExtra)

```

# Example

## Simulate data in wide format

```{r}

# set seed for reproducibility
set.seed(123)

# generate data 
data_wide <- 
  faux::rnorm_multi(n = 100,
                    vars = 5,
                    mu = 3,
                    sd = 1,
                    r = 0.5,
                    varnames = paste0("item_", 1:5),
                    empirical = FALSE) %>%
  rownames_to_column(var = "id")

# recode responses less than 1 or more than 5 to those values, then round scores to whole numbers
# note that {faux} has functions for doing this better
data_wide_likert <- data_wide %>%
  mutate(across(starts_with("item_"), ~ round_half_up(.x, digits = 0))) %>%
  mutate(across(starts_with("item_"), ~ ifelse(.x < 1, 1, ifelse(.x > 5, 5, .x))))

```

## Cronbach's alpha

Wide data like this is a) common and b) useful for calculating metrics like internal consistency.

```{r}

res_alpha <- data_wide_likert %>%
  select(starts_with("item_")) %>%
  psych::alpha()

cronbachs_alpha_estimate <- res_alpha$total$raw_alpha |>
  round_half_up(digits = 2)

```

Cronbach's $\alpha$ = `r cronbachs_alpha_estimate`

## Plot simulated data

```{r}

ggplot(data_wide_likert, aes(x = item_1)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_2)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_3)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_4)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

ggplot(data_wide_likert, aes(x = item_5)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw()

```

- These plots repeat the mortal coding sin of repeating ourselves. If we reshaped the data to 'long' format we could use just one ggplot() call that includes facet_wrap(). 

## Reshape we do this usually with pivot

Using `pivot_longer()`.

```{r}

data_long <- data_wide_likert %>%
  pivot_longer(cols = starts_with("item_"),
               names_to = "item",
               values_to = "response")

ggplot(data_long, aes(x = response)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  theme_linedraw() +
  facet_wrap(~ item)

```

- What other ways could you specify this pivot_longer call's arguments?

### Calculate sum scores

```{r}

sum_scores <- data_long %>%
  group_by(id) %>%
  summarise(sum_score = sum(response))


ggplot(sum_scores, aes(x = sum_score)) +
  geom_histogram(binwidth = 1, boundary = -0.5) +
  scale_x_continuous(breaks = breaks_pretty(n = 10)) +
  theme_linedraw()

```

## Convert this long data back to wide

Just to know how to do it.

```{r}

data_wide_again <- data_long %>%
  pivot_wider(names_from = item,
              values_from = response)

```

## Combine item and sum scores in one data frame

if we would use bindcols we would be fucked

```{r}

data_item_and_sum_scores <- data_wide_again %>%
  left_join(sum_scores, by = "id")

```

# Practice

Wrangle the demographics data included in this exercise more efficiently by reshaping it into wide format. Before, we used filter() to wrangle the age and gender data separately.

here I have to wrangle the data
we did this once before with filter now we try to change the data format from long to wide.

Wide-Format: Mehr Spalten und weniger Zeilen (wenn jede Zeile eine Einheit darstellt).
Long-Format: Weniger Spalten und mehr Zeilen (jede Zeile ist eine Messung).

```{r}
# first I have to load the data
demographics <- read_csv(file = "../data/raw/data_demographics_raw.csv") 

# second I look at the data
show(demographics)

# third I check if the data is in long or in wide format --> Datensatz ist im Long Format
dim(demographics)  # Gibt Zeilen und Spaltenanzahl aus

# Umwandlung von Long- zu Wide-Format --> es entstehen nun die spalten age und gender
demographics_wide <- demographics %>%
  pivot_wider(
    names_from = trialcode,  # Die Spalte, deren Werte neue Spaltennamen werden
    values_from = response   # Die Spalte, deren Werte in die neuen Spalten kommen
  )

# jetzt möchte ich, dass in der alters spalte nur noch numerische werte stehen
demographics_wide <- demographics_wide %>%
  mutate(age = (as.numeric(age)))

# 1 kleiner sanity check: Antworten und Häufigkeiten in der Spalte "age" anzeigen
demographics_wide %>%
  count(age, sort = TRUE)  # Zählt alle Werte und sortiert nach Häufigkeit

# zuerst schaue ich mal was es denn so für Antworten in der Sparte gender hat und lass mir diese anzeigen
demographics_wide %>%
  count(gender, sort = TRUE)  # Zählt alle Werte und sortiert nach Häufigkeit

# jetzt wranglen wir die gender variable
demographics_wide <- demographics_wide %>%
  mutate(
    gender = tolower(gender),  # Standardisierung
    gender = case_when(
      gender %in% c("male", "man", "m") ~ 1,
      gender %in% c("female", "woman", "f") ~ 2,
      gender %in% c("non-binary", "non binary", "nb") ~ 3,
      TRUE ~ NA_real_
    )
  )

# Zeilen entfernen, bei denen sowohl "gender" als auch "age" NA sind
demographics_wide_withoutNAs <- demographics_wide %>%
  filter(!(is.na(gender) & is.na(age)))

# Datensatz auf die gewünschten Spalten reduzieren --> nur noch subject age und gender
demographics_wide_withoutNAs <- demographics_wide_withoutNAs %>%
  select(subject, age, gender)



```

# Session info

```{r}

sessionInfo()

```



